An Efficient Interpolation Technique for Jump Proposals in Reversible-Jump Markov Chain Monte Carlo Calculations

### 2011-03-17 paper.tex
Edits throughout.

Selection among alternative theoretical models given an observed data
set is an important challenge in many areas of physics and astronomy.
In a Bayesian context, model selection involves computing the evidence
for each model given the data.  The model evidence is an integral of
the unnormalized posterior probability distribution over the model
parameter space, representing the probability of obtaining the data set within that model.
Models with larger evidence are preferred; the ratio of the evidences
of two models is the Bayes factor between them.  The product of the
Bayes factor and the ratio of prior probabilities for the two models yields
the odds ratio for the models.
%, the ratio of Bayesian probabilities for the two models given the observation data.

@@

%\ilya{Something about how nested sampling distributes points according to the prior, not the posterior, and therefore can miss modes that are disfavored by prior.  Also, it does not exploit any information about the posterior to make the evidence calculation more efficient.}

Markov Chain Monte Carlo (MCMC) methods attempt to
stochastically produce parameter samples with density proportional to
the posterior probability distribution.   In MCMC techniques, the primary target is an accurate estimate of the posterior distribution.  (We note that an alternative stochastic method for exploring a model parameter space, nested sampling \cite{Skilling:2004, Skilling:2006,Feroz:2009}, focuses on evidence computation rather than sampling the posterior probability density functions.)  \will{Will, are you happy with nested sampling in a parenthetical expression, without further comments?}
It is not straightforward to compute the model evidence from the MCMC
samples.  The most direct way to estimate the evidence for a model
from MCMC samples is to compute the harmonic-mean estimator, but
this estimator of the evidence suffers from infinite
variance\cite{NewtonRaftery:1994,Chib:1995,vanHaasteren:2009}.  Various
methods have been proposed to stabilize the harmonic-mean estimator \will{Citation}, but all
require some assumptions about the shape of the posterior distribution
that may not be satisfied in practice.  MCMC implementations with parallel tempering \cite{EarlDeem:2005} allow for evidence computation via thermodynamic integration, but these can be computationally costly.

@@

Implemented naively, RJMCMC has a significant drawback: because the
chain of samples must be Markovian, only the current sample is
available to the algorithm.  Each time an RJMCMC transitions between
models, the information about the choices of parameter values in the
previous model is lost; subsequent jumps into that model must ``start
fresh,'' and are correspondingly unlikely to be accepted, delaying
convergence of the RJMCMC sample chain.

@@

Reference \cite{Littenberg2009} proposed an alternate method for producing
inter-model jumps in an RJMCMC that relies on a box decomposition of
parameter space, but uses fixed-sized boxes.  The box size is
fixed, so the method cannot adapt to the local structure of the
posterior, and becomes asymptotically inefficient for high-dimensional parameter spaces or highly peaked posteriors.  Meanwhile, the approximation to the
posterior distribution produced by the kD-tree is a constant-in-box
interpolation of the posterior, similar in spirit to the phase-space
density interpolants produced from N-body positions and momenta in
Reference \cite{Ascasibar2005}.  The kD-tree interpolation is effective
in parameter spaces of arbitrary dimensionality, and is quite space-efficient,
requiring $\order{N \log N}$ storage space and $\order{\log N}$ time
to produce each proposed jump, where $N$ is the number of single-model
MCMC samples used to construct the interpolation.

@@

Reversible jump Markov chain Monte Carlo (RJMCMC) \cite{Green1995} is
a technique for Bayesian model comparison.  Below, we give a very
brief introduction to Bayesian analysis; describe a standard MCMC; and
introduce RJMCMC.

@@

When selecting among alternative models, we are interested in finding
the model with the highest posterior probability $p(M_i|d)$.  However,
attempts to directly compute the evidence by performing the
integration in Eq.~\eqref{evidence} are generally very difficult in a
multi-dimensional, multi-modal parameter space when the likelihood has
to be evaluated numerically.  In particular, a grid-based integral
quickly becomes computationally unfeasible as the dimensionality of
$\vtheta$ exceeds a few.  The parameter space must typically be
explored in a stochastic manner before the evidence integral can be
computed.  There are several stochastic parameter-exploration
techniques focused directly on evidence computation (e.g., nested
sampling \cite{Skilling:2004,Skilling:2006} and its variant MultiNest \cite{Feroz:2009}).  Although nested sampling can be used to compute the posterior PDFs within each model along
with the evidences for the various models, perhaps the most common
technique for computing posterior PDFs in the context of a model is
the Markov chain Monte Carlo, which we now describe.

@@

The samples produced by an MCMC algorithm can be used to directly
perform a Monte Carlo evidence integral.  This results in a harmonic
mean estimator for the evidence, which suffers from a large variance
and bias \cite{NewtonRaftery:1994,Chib:1995,vanHaasteren:2009}.
Additional techniques for the direct integration of evidence, also
based on a kD tree decomposition of the parameter space (see
Sec.~\ref{sec:kDTree}), are described in \cite{Weinberg2009}.  These
techniques are promising, but in some cases suffer \cite{Farr2010} \will{Is this really the reference you meant?}
from large variance and bias as well.  An alternative approach to
model selection among a set of models is based on performing an MCMC
in a ``super-model'' that encompasses all of the models under
consideration; this is known the the Reversible Jump Markov chain
Monte Carlo (RJMCMC).

@@

Ref.~\cite{Littenberg2009} decomposed the parameter space into
constant-volume ``bricks'' whose size was set by the typical size of
the peaks of the PDF.  Each sample was associated with the brick that
contained it, and the probability of proposing a jump into a particular brick is thus proportional to the
number of samples within that brick.  Additionally, an extra uniform jump proposal was added to allow for jumps into bricks that did not contain any points, so that the jump 
proposal covers the entire model parameter space.  
%The solution to this problem was to add a uniform distribution to the proposal PDF, in the form of one ``fictitious'' point per brick.  The extra uniform distribution from the fictitious points can represent a significant cost, since inter-model jumps proposed from the uniform distribution are unlikely to be accepted.  
However, the bricks in this
algorithm do not adapt to the local structure of the PDF.  One must
either use small bricks to capture the local structure of the PDF,
placing many bricks in regions without MCMC samples (which can increase memory management and access costs, even for sparse matrices), or use large
bricks, missing the local structure of the PDF in exchange for fewer
empty bricks. \will{Are you happy with this bit of editing?}

@@

The relative error on the determination of the Bayes factor (evidence ratio) scales
with $1/\sqrt{N_{\textnormal{transitions}}}$, where
$N_{\textnormal{transitions}}$ is the number of inter-model
transitions in the RJMCMC.  Thus, as the acceptance rate of
inter-model jumps goes down, the RJMCMC must run longer to achieve a
desired accuracy in the Bayes factor.  By boosting the acceptance
rate of inter-model jumps significantly, the interpolation method
described above can dramatically improve the runtime of an RJMCMC.

@@

As discussed in Section \ref{sec:mcmc}, finding an efficient jump proposal distribution can be a challenge even for a single-model MCMC.  This is particularly difficult when the parameter space is multi-modal, in which case the Markov chain can get stuck on individual high-likelihood islands in the parameter space.  Traversals of low-likelihood oceans by a series of short jumps are extremely unlikely, and random long jumps are almost always rejected, making sampling very inefficient.  Ideally, an analytical understanding of the model that would give a glimpse of the structure of the archipelago of islands would allow for the creation of jump proposal distributions that combine short jumps designed to explore individual islands with directed long jumps that are likely to end up on neighboring islands.  However, it is not always possible to gain such understanding short of performing the MCMC itself.  One solution lies in running multiple chains that explore the parameter space, perhaps getting stuck on local maxima, and then combining the points and interpolating with a kD tree to create a jump proposal distribution.  Although the chains in the first iteration will not have converged, and the relative sizes of the modes will not be accurately known, this will ensure that all of the modes (islands) are at least included in the jump proposal distribution for a second stage, greatly improving its efficiency.  Other modifications to this technique, which we are currently investigating, include multiple iterative stages, with the interpolated samples from each previous stage used as a jump proposal distribution for the next stage, until the posterior PDF is read off from the final stage; and the use of higher temperatures in the early-stage chains to improve sampling.

The possibility of MCMC chains becoming even temporarily stuck on
local maxima also affects convergence.  In the absence of a sufficient
number of jump proposals between the local maxima, the MCMC may not
accurately represent their relative sizes.  For example, this can
happen with antipodal sky location degeneracies for networks of three
ground-based gravitational-wave interferometers or for low-frequency
gravitational wave signals in the proposed space-borne instrument LISA
\cite{LISA}.  Although PDFs produced by MCMC chains will generally
show both locations on the sky, the chain length may not be
sufficiently long in practice to include enough jumps between the two
high-likelihood antipodal locations.  In general, the relative weight
of two separated peaks is determined with a fractional error that
scales as $1/\sqrt{N_{\textnormal{transitions}}}$, where
$N_{\textnormal{transitions}}$ is the number of transitions between
the peaks.  By increasing the number of proposed transitions, the
convergence of such PDFs can be enhanced using the kD tree
interpolated proposal.  We can test the convergence of an MCMC with
local jumps by running a followup MCMC with a jump proposal
distribution from an interpolation of the first MCMC's PDF.  This
follow-up MCMC would make large global jump proposals rather than
small local ones, meaning that it could rapidly accumulate sufficient
points to determine whether the original run spent correct fractional
amounts of time in the various modes.

>>>>

Selection among alternative theoretical models given an observed data
set is an important challenge in many areas of physics and astronomy.
In a Bayesian context, model selection involves computing the evidence
for each model given the data.  The model evidence is an integral of
the unnormalized posterior probability distribution over the model
parameter space, representing the probability of obtaining the data set within that model.
Models with larger evidence are preferred; the ratio of the evidences
of two models is the Bayes factor between them.  The product of the
Bayes factor and the ratio of prior probabilities for the two models yields
the odds ratio for the models.

@@

Markov Chain Monte Carlo (MCMC) methods attempt to stochastically
produce parameter samples with density proportional to the posterior
probability distribution.  In MCMC techniques, the primary target is
an accurate estimate of the posterior distribution.  (We note that an
alternative stochastic method for exploring a model parameter space,
nested sampling \cite{Skilling:2004, Skilling:2006,Feroz:2009},
focuses on evidence computation rather than sampling the posterior
probability density functions.)  It is not straightforward to compute
the model evidence from MCMC samples.  The most direct way to estimate
the evidence for a model from MCMC samples is to compute the
harmonic-mean estimator, but this estimator of the evidence suffers
from infinite
variance\cite{NewtonRaftery:1994,Chib:1995,vanHaasteren:2009}.  MCMC
implementations with parallel tempering \cite{EarlDeem:2005} allow for
evidence computation via thermodynamic integration, but these can be
computationally costly.

@@

Implemented naively, RJMCMC has a significant drawback: because the
chain of samples must be Markovian, only the current sample is
available to the algorithm as it is choosing the next sample.  Each
time an RJMCMC transitions between models, the information about the
choices of parameter values in the previous model is lost; subsequent
jumps into that model must ``start fresh,'' and are correspondingly
unlikely to be accepted, delaying convergence of the RJMCMC sample
chain.

@@

Reference \cite{Littenberg2009} proposed an alternate method for
producing inter-model jumps in an RJMCMC that relies on a box
decomposition of parameter space, but uses fixed-sized boxes.  The box
size is fixed, so the method cannot adapt to the local structure of
the posterior, and becomes asymptotically inefficient for
high-dimensional parameter spaces or highly peaked posteriors.
Meanwhile, the approximation to the posterior distribution produced by
the kD-tree is a constant-in-box interpolation of the posterior,
similar in spirit to the phase-space density interpolants produced
from N-body positions and momenta in Reference \cite{Ascasibar2005}.
The kD-tree interpolation is effective in parameter spaces of
arbitrary dimensionality, and is quite space-efficient, requiring
$\order{N}$ storage space and $\order{\log N}$ time to produce each
proposed jump, where $N$ is the number of single-model MCMC samples
used to construct the interpolation.

@@

Reversible jump Markov chain Monte Carlo (RJMCMC) \cite{Green1995} is
a technique for Bayesian model comparison.  Below, we give a very
brief introduction to Bayesian analysis, describe a standard MCMC, and
introduce RJMCMC.

@@

When selecting among alternative models, we are interested in finding
the model with the highest posterior probability $p(M_i|d)$.  However,
attempts to directly compute the evidence by performing the
integration in Eq.~\eqref{evidence} are generally very difficult in a
multi-dimensional, multi-modal parameter space when the likelihood has
to be evaluated numerically.  In particular, a grid-based integral
quickly becomes computationally unfeasible as the dimensionality of
$\vtheta$ exceeds a few.  The parameter space must typically be
explored in a stochastic manner before the evidence integral can be
computed.  There are several stochastic parameter-exploration
techniques focused directly on evidence computation (e.g., nested
sampling \cite{Skilling:2004,Skilling:2006} and its variant MultiNest
\cite{Feroz:2009}).  Although nested sampling can be used to compute
the posterior PDFs within each model along with the evidences for the
various models, the most common technique for computing posterior PDFs
in the context of a model is the Markov chain Monte Carlo, which we
now describe.

@@

The samples produced by an MCMC algorithm can be used to directly
perform a Monte Carlo evidence integral.  This results in a harmonic
mean estimator for the evidence, which suffers from a large variance
and bias \cite{NewtonRaftery:1994,Chib:1995,vanHaasteren:2009}.
Additional techniques for the direct integration of evidence, also
based on a kD tree decomposition of the parameter space (see
Sec.~\ref{sec:kDTree}), are described in \cite{Weinberg2009}.  These
techniques are promising, but in some cases suffer \cite{Farr2010}
from large variance and bias.  An alternative approach to model
selection among a set of models is based on performing an MCMC in a
``super-model'' that encompasses all of the models under
consideration; this is known the the Reversible Jump Markov chain
Monte Carlo (RJMCMC).

@@

Ref.~\cite{Littenberg2009} decomposed the parameter space into
constant-volume ``bricks'' whose size was set by the typical size of
the peaks of the PDF.  Each sample was associated with the brick that
contained it, and the probability of proposing a jump into a particular brick is thus proportional to the
number of samples within that brick.  Additionally, an extra uniform jump proposal was added to allow for jumps into bricks that did not contain any points, so that the jump 
proposal covers the entire model parameter space.  
However, the bricks in this algorithm do not adapt to the local
structure of the PDF.  One must either use small bricks to capture the
local structure of the PDF, placing many bricks in regions without
MCMC samples (which can increase memory management and access costs),
or use large bricks, missing the local structure of the PDF in
exchange for fewer empty bricks.

@@

The relative error on the determination of the Bayes factor (evidence ratio) scales
with $1/\sqrt{N_{\textnormal{transitions}}}$, where
$N_{\textnormal{transitions}}$ is the number of inter-model
transitions in the RJMCMC.  Thus, as the acceptance rate of
inter-model jumps goes down, the RJMCMC must run longer to achieve a
desired accuracy in the evidence ratio.  By boosting the acceptance
rate of inter-model jumps, the interpolation method described above
can dramatically improve the runtime of an RJMCMC.

@@

As discussed in Section \ref{sec:mcmc}, finding an efficient jump
proposal distribution can be a challenge even for a single-model MCMC.
This is particularly difficult when the parameter space is
multi-modal, in which case the Markov chain can get stuck on
individual high-posterior islands in the parameter space.  Traversals
of low-posterior oceans by a series of short jumps are extremely
unlikely, and random long jumps are almost always rejected, making
sampling very inefficient.  Essentially, such isolated islands in
parameter space behave as separate models, and weighting them by the
number of samples in each peak then corresponds to model selection.
Ideally, an analytical understanding of the model would allow for the
creation of jump proposal distributions that combine short jumps
designed to explore individual islands with directed long jumps that
are likely to end up on neighboring islands.  However, it is not
always possible to gain such understanding short of performing the
MCMC itself.  One solution lies in running multiple chains that
explore the parameter space, perhaps getting stuck on local maxima,
and then combining the points and interpolating with a kD tree to
create a jump proposal distribution.  Although the chains in the first
iteration will not have transitioned between the peaks, and the
samples will therefore not weight the separate peaks appropriately,
this will ensure that all of the modes (islands) are included in the
jump proposal distribution for a second stage.  This second stage can
employ an interpolated jump proposal exactly as in the model selection
discussed above.  Other modifications to this technique, which we are
currently investigating, include multiple iterative stages, with the
interpolated samples from each previous stage used as a jump proposal
distribution for the next stage, until the posterior PDF is read off
from the final stage; and the use of higher temperatures in the
early-stage chains to improve sampling.

An example of a multi-modal parameter space where chains can get stuck
on high-posterior islands comes from antipodal sky location
degeneracies for networks of three ground-based gravitational-wave
interferometers or for low-frequency gravitational wave signals in the
proposed space-borne instrument LISA \cite{LISA}.  Although PDFs
produced by MCMC chains will generally show both locations on the sky,
the chain length may not be sufficiently long in practice to include
enough jumps between the two high-likelihood antipodal locations.  In
general, the relative weight of two separated peaks is determined with
a fractional error that scales as
$1/\sqrt{N_{\textnormal{transitions}}}$, where
$N_{\textnormal{transitions}}$ is the number of transitions between
the peaks.  By increasing the number of proposed transitions, the
convergence of such PDFs can be enhanced using the kD tree
interpolated proposal.
